---
title: "Practical Machine Learning"
author: "Gopinath Rao"
date: "December 22, 2015"
output: html_document
---

#Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#Data
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project is from the this source: http://groupware.les.inf.puc-rio.br/har 

#Task
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. We may use any of the other variables to predict with. A report describing how the model is built must be created, how cross validation is used, what would be the the expected out of sample error, and why the choices is made. The prediction model has to be used to predict 20 different test cases.

#Preliminery Work

##Reproduceability
An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used. Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results below (please see code below for ways and syntax to do so).

##Building the Model
Our outcome variable is classe, a factor variable with 5 levels. For this data set, participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

exactly according to the specification (Class A)

- throwing the elbows to the front (Class B)

- lifting the dumbbell only halfway (Class C)

- lowering the dumbbell only halfway (Class D)

- throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction. Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

###Cross-validation

Cross-validation will be performed by partitioning our training data set randomly without replacement into 2 groups: myTraining data (60% of the original Training data set) and myTesting data (40%). Our models will be fitted on the myTraining data set, and tested on the myTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

###Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the myTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

##Packages and Libraries
The following packages are required for this analysis.
```{r}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)

set.seed(1234)
```

##Getting and Cleaning Data

Read the training and test data into R, identifying "NA", "" and "#DIV/0!" as NA strings
```{r}
training <- read.csv("./pml-training.csv",na.strings=c("NA","","#DIV/0!"))
test <- read.csv("./pml-testing.csv",na.strings=c("NA","","#DIV/0!"))
```

##Cleaning the data

First we want to load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.

```{r}
#delete column of all missing values
training<-training[,colSums(is.na(training)) == 0]
test <-test[,colSums(is.na(test)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
training   <-training[,-c(1:7)]
test <-test[,-c(1:7)]

# and have a look at our new datasets:
dim(training)
dim(test)
```

##Partioning the training data into two

Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

###Observation on the Data

The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the myTraining data set and compare one another.

```{r}
plot(myTraining$classe, col="blue", main="Bar Plot of levels of the variable classe within the myTraining data set", xlab="classe levels", ylab="Frequency")
```

From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 3000 occurrences while level D is the least frequent with about 2000 occurrences.

#Prediction Models

##First Prediction Model : Decision Tree

```{r}
model1 <- rpart(classe ~ ., data=myTraining, method="class")

# Predicting:
prediction1 <- predict(model1, myTesting, type = "class")

# Plot of the Decision Tree
fancyRpartPlot(model1)

# Test results on our myTesting data set:
confusionMatrix(prediction1, myTesting$classe)
```

##Second Prediction Model : Random Forest 



```{r}
model2 <- randomForest(classe ~ ., data=myTraining)

# Predicting:
prediction2 <- predict(model2, myTesting, type = "class")

# Test results on myTesting data set:
confusionMatrix(prediction2, myTesting$classe)
```

##Summary
As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.992 (95% CI: (0.990, 0.994)) compared to 0.739 (95% CI: (0.729, 0.749)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.992. The expected out-of-sample error is estimated at 0.007, or 0.7%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

#Predicting Exercise activity using the model and Assignment Submission function

```{r eval=FALSE}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, test, type="class")

#write file for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```
